---
title: "Cumulative TSS Plot"
author: "elif"
date: "2024-08-14"
output: 
  html_document:
    theme: cerulean
    toc: true
params:
  code_folding: show  # Default setting; user can change this to false to hide code
---

```{r setup, include=FALSE, code_folding=show}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,code_folding = "show" ,collapse = TRUE)
```

# Cumulative TSS Plot

## Table of Contents
{:.no_toc}
* [Calculating the subsampling coverage values for each sample](#calculating-the-subsampling-coverage-values-for-each-sample)

* [Cumulative tss plot for active genes](#cumulative-tss-plot-for-active-genes)

* [Cumulative tss plot for all genes](#cumulative-tss-plot-for-all-genes)


# Calculating the subsampling coverage values for each sample
These coverage values will be used to plot cumulative tss plot


## Loading required packages

```{r}
library(Rsamtools)
library(GenomicAlignments)
library(GenomicRanges)
library(DiffBind)
library(DESeq2)
library(rtracklayer)
library(IRanges)
library(ChIPseeker)
```


## Giving the path of bam files and determining the groups of samples

```{r}
# Define directories and selected samples for the different groups
bam_dirs <- c("data/hdf_hdf", "data/hdf_cd", "data/cd_cd")

selected_samples_hdf_hdf <- c("L132725_sp.bam", "L132727_sp.bam", "L132729_sp.bam", "L132731_sp.bam")
selected_samples_hdf_cd <- c("L132733_sp.bam", "L132735_sp.bam", "L132737_sp.bam", "L132739_sp.bam")
selected_samples_cd_cd <- c("L132741_sp.bam", "L132743_sp.bam", "L132745_sp.bam", "L132747_sp.bam")


```


## Alignment and counting of the reads for all samples

```{r}
# This function is to read a single BAM file, aligned reads and count them.
compute_and_count_reads <- function(bam_file) {
  reads <- readGAlignments(bam_file)
  seqlevelsStyle(reads) <- "Ensembl"
  total_reads <- length(reads)
  return(list(reads = reads, total_reads = total_reads))
}

# To takes a list of directories and samples, iterates over them, and applies the compute_and_count_reads function to each relevant BAM file.


compute_read_counts <- function(bam_dirs, selected_samples_list) {
  read_counts <- list()
  group_names <- c("hdf_hdf", "hdf_cd", "cd_cd")
  
  for (i in seq_along(bam_dirs)) {
    bam_dir <- bam_dirs[i]
    group_name <- group_names[i]
    selected_samples <- selected_samples_list[[i]]
    bam_files <- list.files(bam_dir, pattern = "\\_sp.bam$", full.names = TRUE)
    selected_bam_files <- bam_files[basename(bam_files) %in% selected_samples]
    
    group_read_counts <- sapply(selected_bam_files, function(bam_file) {
      result <- compute_and_count_reads(bam_file)
      total_reads <- result$total_reads
      sample_name <- basename(bam_file)
      print(paste("Library size for sample", sample_name, "in group", group_name, ":", total_reads, "reads"))
      return(total_reads)
    })
    
    read_counts[[group_name]] <- group_read_counts
  }
  
  return(read_counts)
}

```

## Calculate and display result of counting the reads of all groups

```{r}
# To get read counts for all the groups
read_counts <- compute_read_counts(bam_dirs, list(selected_samples_hdf_hdf, selected_samples_hdf_cd, selected_samples_cd_cd))

# Find the smallest number of reads across all samples
min_reads <- min(unlist(read_counts))

print(paste("Minimum reads for subsampling:", min_reads))
```

## Subsampling the reads of samples by using minimum reads count

```{r}
# Function to take reads and return subsample of them
subsample_reads <- function(reads, num_reads) {
  if (length(reads) > num_reads) {
    set.seed(123)  # For reproducibility
    sampled_reads <- sample(reads, num_reads)
  } else {
    sampled_reads <- reads
  }
  return(sampled_reads)
}


#Selected samples are determined to prevent error that is taken from compute_subsampled_coverage function
selected_samples_list <- list(
  hdf_hdf = selected_samples_hdf_hdf,
  hdf_cd = selected_samples_hdf_cd,
  cd_cd = selected_samples_cd_cd
)


#Function to subsample the reads and then calculates the coverage on the subsampled reads
compute_subsampled_coverage <- function(bam_dirs, selected_samples_list, num_reads) {
  
  # Initialize coverage_list with the correct group names
  coverage_list <- as.list(rep(NA, length(selected_samples_list)))
  if (!is.null(names(selected_samples_list))) {
    names(coverage_list) <- names(selected_samples_list)
  } else {
    names(coverage_list) <- paste0("group_", seq_along(selected_samples_list))
  }
  
  for (i in seq_along(bam_dirs)) {
    bam_dir <- bam_dirs[i]
    group_name <- names(coverage_list)[i]
    selected_samples <- selected_samples_list[[i]]
    bam_files <- list.files(bam_dir, pattern = "\\_sp.bam$", full.names = TRUE)
    selected_bam_files <- bam_files[basename(bam_files) %in% selected_samples]
    
    group_coverage_list <- lapply(selected_bam_files, function(bam_file) {
      result <- compute_and_count_reads(bam_file)
      subsampled_reads <- subsample_reads(result$reads, num_reads)
      gr_subsampled <- granges(subsampled_reads)
      coverage <- coverage(gr_subsampled)
      return(coverage)
    })
    
    names(group_coverage_list) <- basename(selected_bam_files)
    
    # Assign group_coverage_list to coverage_list[[group_name]]
    coverage_list[[group_name]] <- group_coverage_list
  }
  return(coverage_list)
}

#Computing coverage by using subsampled reads with compute_subsampled_coverage function for all selected samples
coverage_list_subsampled <- compute_subsampled_coverage(bam_dirs, selected_samples_list, min_reads)
```

# Cumulative tss plot for active genes

## Loading the necessary libraries 

```{r}
# Load necessary libraries
library(tsTools)
library(IRanges)
library(org.Mm.eg.db)
library(GenomicRanges)
library(SummarizedExperiment)
library(GenomicAlignments)
library(rtracklayer)
library(TxDb.Mmusculus.UCSC.mm39.refGene)
```

## Reading the summarizedExperiment object to get active genes

```{r}
# Load the summarized experiment object
se <- readRDS("data/se.rds")

# View the structure of the SummarizedExperiment object
str(se)

```

## Getting the chromosome information 

```{r}
# Check if elementMetadata has the chromosome information,it display more information about the active genes 
element_metadata <- elementMetadata(se)

str(element_metadata)

```

## Calculating the expressions for all genes and determining the treshold for obtaining active genes 

```{r}
# Access the expression data from the SummarizedExperiment object
expression_data <- assay(se, "counts") #It could be also tpm but here we have count

# Calculate the average expression for each gene across all samples
average_expression <- rowMeans(expression_data)

activity_threshold <- 1

# Identify genes with average expression above the threshold
active_gene_indices <- which(average_expression > activity_threshold)

# Ensure that element_metadata has gene IDs and potentially chromosome information
if (!is.null(element_metadata)) {
  # Filter metadata to include only the active genes
  active_gene_ids <- rownames(expression_data)[active_gene_indices]
  active_genes_metadata <- element_metadata[rownames(element_metadata) %in% active_gene_ids, ]

  # Optionally, merge with expression data for a complete view
  active_genes <- cbind(expression_data[active_gene_ids, ], active_genes_metadata)

  # Display the number of active genes
  print(length(active_gene_ids))

  # Optionally, inspect the active genes
  head(active_genes)
} else {
  print("No element metadata available.")
}

```

```{r}

# Getting the gene annotations from database
ann <- genes(TxDb.Mmusculus.UCSC.mm39.refGene)

seqlevelsStyle(ann)<-"ensembl"

#Names in ann object refers to entrez ID of the gene which is in active_genes_metadata object
names(ann)

# Filter the `ann` object to include only active genes. While names(ann) refers to the entrez ID of the gene, i mapped entrez id to ensembl id 
ann_active <- ann[names(ann) %in% element_metadata$entrez]

# Modify the genomic ranges of the genes to focus on the TSS by adjusting the gene's range to a single point "start"
ann_active <- resize(ann_active, 0, fix="start")

# Assign names to each gene in the `ann_active` object
names(ann_active) <- paste("gene", 1:length(ann_active), sep="_")


```

## Cumulative tss plot for treatment group by active genes 

```{r}
# Set up the plotting area to have 2 rows and 2 columns
par(mfrow = c(2, 2))

# Determine the samples in treatment group to get their coverage values from coverage list
selected_samples_hdf_hdf <- c("L132725_sp.bam", "L132727_sp.bam", "L132729_sp.bam", "L132731_sp.bam")
samples_hdf_hdf <- selected_samples_hdf_hdf

#Create empty list to store y axis values of each sample.Then we will find the maximum one and use it to scale y axis. In this way each sample will be in same scale
all_results <- list()

# Iterate through each sample in the hdf_hdf group
for (sample in samples_hdf_hdf) {
  cat("Processing sample:", sample, "\n")
  
  # Calculate coverage windows for the current sample
  result <- coverageWindowsCenteredStranded(ann_active, window.size = 2000, coverage_list_subsampled$hdf_hdf[[sample]])
  
  # Store the result
  all_results[[sample]] <- apply(result, 2, mean)
}

# Determine the maximum y-value across all results to use as max y value
y_max <- max(sapply(all_results, max))
  

 # Plot each sample with the determined y-limit
for (sample in samples_hdf_hdf) {
  
  x <- seq(-1000, 1000)
  
  # Plot the cumulative TSS for the current sample
  plot(x, all_results[[sample]], type = "l", 
       xlab = "Position relative to center position", ylab = "Dyad density",
       main = paste("Cumulative TSS for sample:", sample),
       ylim = c(0, y_max))
}

# Reset the plotting layout to default
par(mfrow = c(1, 1)) 


```
## Cumulative tss plot for control group by active genes 

```{r}
# Set up the plotting area to have 2 rows and 2 columns
par(mfrow = c(2, 2))

#Determining the samples in control group to get their coverage values from coverage list 
selected_samples_cd_cd <- c("L132741_sp.bam", "L132743_sp.bam", "L132745_sp.bam", "L132747_sp.bam")
samples_cd_cd <- selected_samples_cd_cd

all_results <- list()

# Iterate through each sample in the cd_cd group
for (sample in samples_cd_cd) {
  cat("Processing sample:", sample, "\n")
  
  # Calculate coverage windows for the current sample
  result <- coverageWindowsCenteredStranded(ann_active, window.size = 2000, coverage_list_subsampled$cd_cd[[sample]])
  
  # Store the result
  all_results[[sample]] <- apply(result, 2, mean)
}
# Determine the maximum y-value across all results
y_max <- max(sapply(all_results, max))
  

 # Plot each sample with the determined y-limit
for (sample in samples_cd_cd) {
  
  x <- seq(-1000, 1000)
  
  # Plot the cumulative TSS for the current sample
  plot(x, all_results[[sample]], type = "l", 
       xlab = "Position relative to center position", ylab = "Dyad density",
       main = paste("Cumulative TSS for sample:", sample),
       ylim = c(0, y_max))
}

# Reset the plotting layout to default
par(mfrow = c(1, 1)) 


```


## Cumulative tss plot for reverse group by active genes 

```{r}
# Set up the plotting area to have 2 rows and 2 columns
par(mfrow = c(2, 2))


#Determining the samples in reverse group to get their coverage values from coverage list 
selected_samples_hdf_cd <- c("L132733_sp.bam", "L132735_sp.bam", "L132737_sp.bam", "L132739_sp.bam")
samples_hdf_cd <- selected_samples_hdf_cd

all_results <- list()

# Iterate through each sample in the hdf_cd group
for (sample in samples_hdf_cd) {
  cat("Processing sample:", sample, "\n")
  
  # Calculate coverage windows for the current sample
  result <- coverageWindowsCenteredStranded(ann_active, window.size = 2000, coverage_list_subsampled$hdf_cd[[sample]])
  
  # Store the result
  all_results[[sample]] <- apply(result, 2, mean)
}
# Determine the maximum y-value across all results
y_max <- max(sapply(all_results, max))
  

 # Plot each sample with the determined y-limit
for (sample in samples_hdf_cd) {
  
  x <- seq(-1000, 1000)
  
  # Plot the cumulative TSS for the current sample
  plot(x, all_results[[sample]], type = "l", 
       xlab = "Position relative to center position", ylab = "Dyad density",
       main = paste("Cumulative TSS for sample:", sample),
       ylim = c(0, y_max))
}

# Reset the plotting layout to default
par(mfrow = c(1, 1)) 


```


# Cumulative tss plot for all genes 

## Getting promoters of all genes by using database

```{r}
#To get the promoters from annotation file we can use also getPromoters() function ann<-getPromoters(TxDb.Mmusculus.UCSC.mm39.refGene,by="gene")

# Getting the gene annotations from database
ann <- genes(TxDb.Mmusculus.UCSC.mm39.refGene)

# While coverage list and ann object have different chromosome naming style, make them same
seqlevelsStyle(ann) <- "ensembl"

# Modifying genomic ranges of the genes to focus the tss by adjusting the gene's range as a single point "start" which is tss of gene
ann <- resize(ann, 0, fix="start")

# Assigning names to each gene in ann object like gene_1
names(ann) <- paste("gene", 1:length(ann), sep="_")



```

## Cumulative tss plot for treatment group

```{r}

# Set up the plotting area to have 2 rows and 2 columns
par(mfrow = c(2, 2))

# Determine the samples in treatment group to get their coverage values from coverage list
selected_samples_hdf_hdf <- c("L132725_sp.bam", "L132727_sp.bam", "L132729_sp.bam", "L132731_sp.bam")
samples_hdf_hdf <- selected_samples_hdf_hdf
all_results <- list()

# Iterate through each sample in the hdf_hdf group
for (sample in samples_hdf_hdf) {
  cat("Processing sample:", sample, "\n")
  
  # Calculate coverage windows for the current sample
  result <- coverageWindowsCenteredStranded(ann, window.size = 2000, coverage_list_subsampled$hdf_hdf[[sample]])
  
  # Store the result
  all_results[[sample]] <- apply(result, 2, mean)
}
# Determine the maximum y-value across all results
y_max <- max(sapply(all_results, max))
  

 # Plot each sample with the determined y-limit
for (sample in samples_hdf_hdf) {
  
  x <- seq(-1000, 1000)
  
  # Plot the cumulative TSS for the current sample
  plot(x, all_results[[sample]], type = "l", 
       xlab = "Position relative to center position", ylab = "Dyad density",
       main = paste("Cumulative TSS for sample:", sample),
       ylim = c(0, y_max))
}

# Reset the plotting layout to default
par(mfrow = c(1, 1)) 


```

## Cumulatuve tss plot for control group

```{r}

# Set up the plotting area to have 2 rows and 2 columns
par(mfrow = c(2, 2))

#Determining the samples in control group to get their coverage values from coverage list 
selected_samples_cd_cd <- c("L132741_sp.bam", "L132743_sp.bam", "L132745_sp.bam", "L132747_sp.bam")
samples_cd_cd <- selected_samples_cd_cd
all_results <- list()

# Iterate through each sample in the cd_cd group
for (sample in samples_cd_cd) {
  cat("Processing sample:", sample, "\n")
  
  # Calculate coverage windows for the current sample
  result <- coverageWindowsCenteredStranded(ann, window.size = 2000, coverage_list_subsampled$cd_cd[[sample]])
  
  # Store the result
  all_results[[sample]] <- apply(result, 2, mean)
}
# Determine the maximum y-value across all results
y_max <- max(sapply(all_results, max))
  

 # Plot each sample with the determined y-limit
for (sample in samples_cd_cd) {
  
  x <- seq(-1000, 1000)
  
  # Plot the cumulative TSS for the current sample
  plot(x, all_results[[sample]], type = "l", 
       xlab = "Position relative to center position", ylab = "Dyad density",
       main = paste("Cumulative TSS for sample:", sample),
       ylim = c(0, y_max))
}

# Reset the plotting layout to default
par(mfrow = c(1, 1)) 



```


## Cumulative tss plot for reverse group

```{r}
# Set up the plotting area to have 2 rows and 2 columns
par(mfrow = c(2, 2))


#Determining the samples in reverse group to get their coverage values from coverage list 
selected_samples_hdf_cd <- c("L132733_sp.bam", "L132735_sp.bam", "L132737_sp.bam", "L132739_sp.bam")
samples_hdf_cd <- selected_samples_hdf_cd
all_results <- list()

# Iterate through each sample in the hdf_cd group
for (sample in samples_hdf_cd) {
  cat("Processing sample:", sample, "\n")
  
  # Calculate coverage windows for the current sample
  result <- coverageWindowsCenteredStranded(ann, window.size = 2000, coverage_list_subsampled$hdf_cd[[sample]])
  
  # Store the result
  all_results[[sample]] <- apply(result, 2, mean)
}
# Determine the maximum y-value across all results
y_max <- max(sapply(all_results, max))
  
#If we want to display result of first sample 
all_results[[1]]

 # Plot each sample with the determined y-limit
for (sample in samples_hdf_cd) {
  
  x <- seq(-1000, 1000)
  
  # Plot the cumulative TSS for the current sample
  plot(x, all_results[[sample]], type = "l", 
       xlab = "Position relative to center position", ylab = "Dyad density",
       main = paste("Cumulative TSS for sample:", sample),
       ylim = c(0, y_max))
}

# Reset the plotting layout to default
par(mfrow = c(1, 1)) 


```


