---
title: "Peak Visualization"
author: "elif"
date: "2024-07-24"
output: 
  html_document: 
    theme: cerulean
    toc: true
params:
  code_folding: show  # Default setting; user can change this to false to hide code
---
```{r setup, include=FALSE, code_folding=show}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,code_folding = "show" ,collapse = TRUE)

```

# Top Peak Visualization

## Table of Contents
{:.no_toc}
* [Loading required packages](#loading-required-packages)

* [Reading Bam files of each sample and converting them to coverage](#reading-bam-files-of-each-sample-and-converting-them-to-coverage)

* [Selecting samples from each group](#selecting-samples-from-each-group)

* [Top peak of Treatment and Control group comparison](#top-peak-of-treatment-and-control-group-comparison)

* [Top peak of Reverse and Control group comparison](#top-peak-of-reverse-and-control-group-comparison)

* [Top peak of Treatment and Reverse group comparison](#top-peak-of-treatment-and-reverse-group-comparison)

* [Annotation for top peaks](#annotation-for-top-peaks)

* [View Annotation Results](#view-annotation-results)

* [Comparison of Treatment and Reverse Group In Same Plot](#comparison-of-treatment-and-reverse-group-in-same-plot)

* [Comparison of Treatment and Control Group In Same Plot](#comparison-of-treatment-and-control-group-in-same-plot)

* [Comparison of Control and Reverse Group In Same Plot](#comparison-of-control-and-reverse-group-in-same-plot)

## Loading required packages
```{r}
library(Rsamtools)
library(GenomicAlignments)
library(GenomicRanges)
library(DiffBind)
library(DESeq2)
library(readr)
library(tsTools)
require(TxDb.Mmusculus.UCSC.mm39.refGene)
library(RColorBrewer)
library(GEOquery)
library(rtracklayer)
library(IRanges)
library(ggplot2)
library(ChIPseeker)
library(org.Mm.eg.db)
```


```{r}
# Define directories for BAM files
bam_dirs <- c("data/hdf_hdf", "data/hdf_cd", "data/cd_cd")

# Function to read BAM files and compute coverage
compute_coverage <- function(bam_dir) {
  bam_files <- list.files(bam_dir, pattern = "\\_sp.bam$", full.names = TRUE)
  coverage_list <- lapply(bam_files, function(bam_file) {
    reads <- readGAlignments(bam_file)
    read_count <- length(reads)
    coverage <- coverage(granges(reads))
    return(coverage)
  })
 
  
# Normalize each sample's coverage by its own minimum read count, handling the case when min read count is 0
  normalized_coverage_list <- lapply(coverage_list, function(cov) {
    min_read_count <- min(cov)
    if (min_read_count == 0) {
      return(cov)  # Return the original coverage if min read count is 0
    } else {
      return(cov / min_read_count)
    }
  })
  
  names(normalized_coverage_list) <- basename(bam_files)  # Set names for each coverage object same with sample name
  return(normalized_coverage_list)
}


# Compute coverage for each group
coverage_list_hdf_hdf <- compute_coverage(bam_dirs[1])
coverage_list_hdf_cd <- compute_coverage(bam_dirs[2])
coverage_list_cd_cd <- compute_coverage(bam_dirs[3])



create_simple_rle_list <- function(coverage_list) {
  cov <- SimpleList() #when i try as SimpleRleList() it gave error function cannot found
  for (sample_name in names(coverage_list)) {
    cov[[sample_name]] <- coverage_list[[sample_name]]
  }
  return(cov)
}

cov_hdf_hdf <- create_simple_rle_list(coverage_list_hdf_hdf)
cov_hdf_cd <- create_simple_rle_list(coverage_list_hdf_cd)
cov_cd_cd <- create_simple_rle_list(coverage_list_cd_cd)

```




## Reading Bam files of each sample and converting them to coverage
```{r}

# Define directories for BAM files
bam_dirs <- c("data/hdf_hdf", "data/hdf_cd", "data/cd_cd")

# Function to read BAM files and compute coverage
compute_coverage <- function(bam_dir) {
  bam_files <- list.files(bam_dir, pattern = "\\_sp.bam$", full.names = TRUE)
  coverage_list <- lapply(bam_files, function(bam_file) {
    reads <- readGAlignments(bam_file)
    coverage <- coverage(granges(reads))
    return(coverage)
  })
  names(coverage_list) <- basename(bam_files)  # Set names for each coverage object same with sample name
  return(coverage_list)
}

# Compute coverage for each group
coverage_list_hdf_hdf <- compute_coverage(bam_dirs[1])
coverage_list_hdf_cd <- compute_coverage(bam_dirs[2])
coverage_list_cd_cd <- compute_coverage(bam_dirs[3])



create_simple_rle_list <- function(coverage_list) {
  cov <- SimpleList() #when i try as SimpleRleList() it gave error function cannot found
  for (sample_name in names(coverage_list)) {
    cov[[sample_name]] <- coverage_list[[sample_name]]
  }
  return(cov)
}

cov_hdf_hdf <- create_simple_rle_list(coverage_list_hdf_hdf)
cov_hdf_cd <- create_simple_rle_list(coverage_list_hdf_cd)
cov_cd_cd <- create_simple_rle_list(coverage_list_cd_cd)


```

## Selecting samples from each group
```{r}
# Define the samples you want to visualize for each group
selected_samples_hdf_hdf <- c("L132725_sp.bam", "L132727_sp.bam","L132729_sp.bam","L132731_sp.bam") 
selected_samples_hdf_cd <- c("L132733_sp.bam", "L132735_sp.bam","L132737_sp.bam","L132739_sp.bam")   
selected_samples_cd_cd <- c("L132741_sp.bam", "L132743_sp.bam","L132745_sp.bam","L132747_sp.bam")    

# Function to create a profile list for selected samples
create_profile_list <- function(cov, sample_names) {
  profs <- cov[sample_names]
  return(profs)
}

# Create profile lists for selected samples
profs_hdf_hdf <- create_profile_list(cov_hdf_hdf, selected_samples_hdf_hdf)
profs_hdf_cd <- create_profile_list(cov_hdf_cd, selected_samples_hdf_cd)
profs_cd_cd <- create_profile_list(cov_cd_cd, selected_samples_cd_cd)



```

## Top peak of Treatment and Control group comparison
```{r}
#Determining profs, samples that are visualized, outside the plotting code is preferred.
profs1 = list("sample1" = profs_hdf_hdf[[selected_samples_hdf_hdf[1]]],
                         "sample2" = profs_hdf_hdf[[selected_samples_hdf_hdf[2]]],
                         "sample3" = profs_hdf_hdf[[selected_samples_hdf_hdf[3]]],
                         "sample4" = profs_hdf_hdf[[selected_samples_hdf_hdf[4]]]
             )
#To match chr parameter with coverages, two of them will be ensembl type
seqlevelsStyle(TxDb.Mmusculus.UCSC.mm39.refGene) <- "ensembl"

#To indicate actual place of top peak
ann <- list(
  "sample4" = data.frame(chr = "6", start = 5510146, end = 5510546, col=c("blue")))

# Plot profiles for a specific region and chromosome
#for fstart and fend i substract/add 2kb to actual top peak range
plot1<-plotProfiles(fstart = 5508146, fend = 5517546, fchr = "6",
             profs = profs1, ylims=list(c(0,200),c(0,200),c(0,200),c(0,200)),
             cols = brewer.pal(length(selected_samples_hdf_hdf), "Set1"),
             txdb = TxDb.Mmusculus.UCSC.mm39.refGene,
             ann=ann)
            
            
```

```{r}

#To indicate actual place of top peak
ann <- list(
  "sample4" = data.frame(chr = "6", start = 5510146, end = 5510546, col=c("blue")))

profs2 = list("sample1" = profs_cd_cd[[selected_samples_cd_cd[1]]],
                         "sample2" = profs_cd_cd[[selected_samples_cd_cd[2]]],
                         "sample3" = profs_cd_cd[[selected_samples_cd_cd[3]]],
                         "sample4" = profs_cd_cd[[selected_samples_cd_cd[4]]]
             )

# Plot profiles for a specific region and chromosome
#for fstart and fend i substract/add 2kb to actual top peak range
plot2<-plotProfiles(fstart = 5508146, fend = 5517546, fchr = "6",
             profs = profs2,ylims=list(c(0,200),c(0,200),c(0,200),c(0,200)),
             cols = brewer.pal(length(selected_samples_cd_cd), "Set1"),
             txdb = TxDb.Mmusculus.UCSC.mm39.refGene,
             ann=ann)
```


## Top peak of Reverse and Control group comparison
```{r}
#To indicate actual place of top peak
ann <- list(
  "sample4" = data.frame(chr = "15", start = 76330515, end = 76330915, col=c("blue")))
profs2 = list("sample1" = profs_cd_cd[[selected_samples_cd_cd[1]]],
                         "sample2" = profs_cd_cd[[selected_samples_cd_cd[2]]],
                         "sample3" = profs_cd_cd[[selected_samples_cd_cd[3]]],
                         "sample4" = profs_cd_cd[[selected_samples_cd_cd[4]]]
             )

# Plot profiles for a specific region and chromosome
#for fstart and fend i substract/add 2kb to actual top peak range
plot3<-plotProfiles(fstart = 76328515, fend = 76332915, fchr = "15",
             profs = profs2,ylims=list(c(0,200),c(0,200),c(0,200),c(0,200)),
             cols = brewer.pal(length(selected_samples_cd_cd), "Set1"),
             txdb = TxDb.Mmusculus.UCSC.mm39.refGene,
             ann=ann)


```

```{r}
#To indicate actual place of top peak
ann <- list(
  "sample4" = data.frame(chr = "15", start = 76330515, end = 76330915, col=c("blue")))
profs3 = list("sample1" = profs_hdf_cd[[selected_samples_hdf_cd[1]]],
                         "sample2" = profs_hdf_cd[[selected_samples_hdf_cd[2]]],
                         "sample3" = profs_hdf_cd[[selected_samples_hdf_cd[3]]],
                         "sample4" = profs_hdf_cd[[selected_samples_hdf_cd[4]]]
             )

# Plot profiles for a specific region and chromosome
#for fstart and fend i substract/add 2kb to actual top peak range
plot4<-plotProfiles(fstart = 76328515, fend = 76332915, fchr = "15",
             profs = profs3,ylims=list(c(0,200),c(0,200),c(0,200),c(0,200)),
             cols = brewer.pal(length(selected_samples_hdf_cd), "Set1"),
             txdb = TxDb.Mmusculus.UCSC.mm39.refGene,
             ann=ann)
```


## Top peak of Treatment and Reverse group comparison

```{r}
#To indicate actual place of top peak
ann <- list(
  "sample4" = data.frame(chr = "13", start = 22219161, end = 22219561, col=c("blue")))

profs1 = list("sample1" = profs_hdf_hdf[[selected_samples_hdf_hdf[1]]],
                         "sample2" = profs_hdf_hdf[[selected_samples_hdf_hdf[2]]],
                         "sample3" = profs_hdf_hdf[[selected_samples_hdf_hdf[3]]],
                         "sample4" = profs_hdf_hdf[[selected_samples_hdf_hdf[4]]]
             )

# Plot profiles for a specific region and chromosome
#for fstart and fend i substract/add 2kb to actual top peak range
plot5<-plotProfiles(fstart = 22217161, fend = 22221561, fchr = "13",
             profs = profs1,ylims=list(c(0,200),c(0,200),c(0,200),c(0,200)),
             cols = brewer.pal(length(selected_samples_hdf_hdf), "Set1"),
             txdb = TxDb.Mmusculus.UCSC.mm39.refGene,
             ann=ann)



```

```{r}
ann <- list(
  "sample4" = data.frame(chr = "13", start = 22219161, end = 22219561, col=c("blue")))

profs3= list("sample1" = profs_hdf_cd[[selected_samples_hdf_cd[1]]],
                         "sample2" = profs_hdf_cd[[selected_samples_hdf_cd[2]]],
                         "sample3" = profs_hdf_cd[[selected_samples_hdf_cd[3]]],
                         "sample4" = profs_hdf_cd[[selected_samples_hdf_cd[4]]]
             )

# Plot profiles for a specific region and chromosome
#for fstart and fend i substract/add 2kb to actual top peak range
plot6<-plotProfiles(fstart = 22217161, fend = 22221561, fchr = "13",
             profs = profs3,ylims=list(c(0,200),c(0,200),c(0,200),c(0,200)),
             cols = brewer.pal(length(selected_samples_hdf_cd), "Set1"),
             txdb = TxDb.Mmusculus.UCSC.mm39.refGene,
             ann=ann)
```

## Annotation for top peaks

```{r}
# Example data for top peaks
top_peak1 <- data.frame(
    seqnames = "6",
    start = 5508146,
    end = 5517546
)
top_peak2 <- data.frame(
    seqnames = "15",
    start = 76328515,
    end = 76332915
)
top_peak3 <- data.frame(
    seqnames = "13",
    start = 22219161,
    end = 22219561
)

# Create GRanges objects
gr_peak1 <- GRanges(seqnames=top_peak1$seqnames, ranges=IRanges(start=top_peak1$start, end=top_peak1$end))
gr_peak2 <- GRanges(seqnames=top_peak2$seqnames, ranges=IRanges(start=top_peak2$start, end=top_peak2$end))
gr_peak3 <- GRanges(seqnames=top_peak3$seqnames, ranges=IRanges(start=top_peak3$start, end=top_peak3$end))

txdb <- TxDb.Mmusculus.UCSC.mm39.refGene  

# Annotate peaks
peakAnno1 <- annotatePeak(gr_peak1, tssRegion=c(-3000, 3000), TxDb=txdb, annoDb="org.Mm.eg.db")
peakAnno2 <- annotatePeak(gr_peak2, tssRegion=c(-3000, 3000), TxDb=txdb, annoDb="org.Mm.eg.db")
peakAnno3 <- annotatePeak(gr_peak3, tssRegion=c(-3000, 3000), TxDb=txdb, annoDb="org.Mm.eg.db")


```

## View Annotation Results

```{r}
# Convert to data frames
df_peakAnno1 <- as.data.frame(peakAnno1)
df_peakAnno2 <- as.data.frame(peakAnno2)
df_peakAnno3 <- as.data.frame(peakAnno3)

# View annotation results
head(df_peakAnno1)
head(df_peakAnno2)
head(df_peakAnno3)
```


## Comparison of Treatment and Reverse Group In Same Plot

```{r}

#To match chr parameter with coverages, two of them will be ensembl type
seqlevelsStyle(TxDb.Mmusculus.UCSC.mm39.refGene) <- "ensembl"

ann <- list(
  "sample4" = data.frame(chr = "13", start = 22219161, end = 22219561, col=c("green")))

profs3_1= list("sample1_reverse" = profs_hdf_cd[[selected_samples_hdf_cd[1]]],
                         "sample2" = profs_hdf_cd[[selected_samples_hdf_cd[2]]],
                         "sample3" = profs_hdf_cd[[selected_samples_hdf_cd[3]]],
                         "sample4" = profs_hdf_cd[[selected_samples_hdf_cd[4]]],
                         "sample1_treatment" = profs_hdf_hdf[[selected_samples_hdf_hdf[1]]],
                         "sample2" = profs_hdf_hdf[[selected_samples_hdf_hdf[2]]],
                         "sample3" = profs_hdf_hdf[[selected_samples_hdf_hdf[3]]],
                         "sample4" = profs_hdf_hdf[[selected_samples_hdf_hdf[4]]]
             )

# Generate two colors from the Set1 palette
colors <- brewer.pal(2, "Set1")

# Create a vector of colors: first four samples one color, next four samples another color
profs_colors <- c(rep(colors[1], 4), rep(colors[2], 4))

plot7<-plotProfiles(fstart = 22217161, fend = 22221561, fchr = "13",
             profs = profs3_1, ylims=list(c(0,200),c(0,200),c(0,200),c(0,200),c(0,200),c(0,200),c(0,200),c(0,200)),
             cols = profs_colors,  # Use the custom colors for each sample
             txdb = TxDb.Mmusculus.UCSC.mm39.refGene,
             ann=ann)
```

## Comparison of Treatment and Control Group In Same Plot

```{r}
#Determining profs, samples that are visualized, outside the plotting code is preferred.
profs1_2 = list("sample1_treatment" = profs_hdf_hdf[[selected_samples_hdf_hdf[1]]],
                         "sample2" = profs_hdf_hdf[[selected_samples_hdf_hdf[2]]],
                         "sample3" = profs_hdf_hdf[[selected_samples_hdf_hdf[3]]],
                         "sample4" = profs_hdf_hdf[[selected_samples_hdf_hdf[4]]],
                         "sample1_control" = profs_cd_cd[[selected_samples_cd_cd[1]]],
                         "sample2" = profs_cd_cd[[selected_samples_cd_cd[2]]],
                         "sample3" = profs_cd_cd[[selected_samples_cd_cd[3]]],
                         "sample4" = profs_cd_cd[[selected_samples_cd_cd[4]]]
             )

# Generate two colors from the Set1 palette
colors <- brewer.pal(2, "Set1")

# Create a vector of colors: first four samples one color, next four samples another color
profs_colors <- c(rep(colors[1], 4), rep(colors[2], 4))

#To indicate actual place of top peak
ann <- list(
  "sample4" = data.frame(chr = "6", start = 5510146, end = 5510546, col=c("green")))

# Plot profiles for a specific region and chromosome
#for fstart and fend i substract/add 2kb to actual top peak range
plot1<-plotProfiles(fstart = 5508146, fend = 5517546, fchr = "6",
             profs = profs1_2, ylims=list(c(0,200),c(0,200),c(0,200),c(0,200),c(0,200),c(0,200),c(0,200),c(0,200)),
             cols = profs_colors,
             txdb = TxDb.Mmusculus.UCSC.mm39.refGene,
             ann=ann)
```


## Comparison of Control and Reverse Group In Same Plot

```{r}
# Generate two colors from the Set1 palette
colors <- brewer.pal(2, "Set1")

# Create a vector of colors: first four samples one color, next four samples another color
profs_colors <- c(rep(colors[1], 4), rep(colors[2], 4))

#To indicate actual place of top peak
ann <- list(
  "sample4" = data.frame(chr = "15", start = 76330515, end = 76330915, col=c("green")))

profs2_3 = list("sample1_control" = profs_cd_cd[[selected_samples_cd_cd[1]]],
                         "sample2" = profs_cd_cd[[selected_samples_cd_cd[2]]],
                         "sample3" = profs_cd_cd[[selected_samples_cd_cd[3]]],
                         "sample4" = profs_cd_cd[[selected_samples_cd_cd[4]]],
                         "sample1_reverse" = profs_hdf_cd[[selected_samples_hdf_cd[1]]],
                         "sample2" = profs_hdf_cd[[selected_samples_hdf_cd[2]]],
                         "sample3" = profs_hdf_cd[[selected_samples_hdf_cd[3]]],
                         "sample4" = profs_hdf_cd[[selected_samples_hdf_cd[4]]]
             )

# Plot profiles for a specific region and chromosome
#for fstart and fend i substract/add 2kb to actual top peak range
plot3<-plotProfiles(fstart = 76328515, fend = 76332915, fchr = "15",
             profs = profs2_3, ylims=list(c(0,200),c(0,200),c(0,200),c(0,200),c(0,200),c(0,200),c(0,200),c(0,200)),
             cols = profs_colors,
             txdb = TxDb.Mmusculus.UCSC.mm39.refGene,
             ann=ann)
```

