---
title: "task1_final"
author: "elif"
date: "2024-12-16"
output: 
  html_document:
    theme: cerulean
    toc: true
params:
  code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# PCA and UMAP Using Seurat Object of One Sample

## Table of Contents
{:.no_toc}

* [Load neccessary packages and call libraries](#load-necessary-packages-and-call-libraries)

* [Import expression data and create seurat object](#import-expression-data-and-create-seurat-objects)

* [Normalize the matrix find highly variable genes and scale the data](#normalize-the-matrix-find-highly-variable-genes-and-scale-the-data)

* [PCA plot](#pca-plot)

* [Cluster the cells and perform UMAP ](#cluster-the-cells-and-perform-umap)

##Load neccessary packages and call libraries

```{r }
library(Seurat)
library(ggplot2)
```

##Import expression data and create seurat object

For six sample from 2 patient. One primary two lymph node samples from each patient. These samples were found in GSE180286_RAW folder, from second file to seventh file. 

```{r}

# Set the working directory
setwd("/Users/elifkarahaner/Desktop/scrna/")

# Get the list of files in the directory
file_paths <- list.files("GSE180286_RAW", full.names = TRUE)[2:7]

# Initialize an empty list to store Seurat objects
seurat_objects <- list()
seurat_objects[1]
# Loop over each file and create a Seurat object
for (file_path in file_paths) {
  # Load the data
  data <- read.table(gzfile(file_path), header = TRUE, sep = "\t")
  
  # Get a clean name for the Seurat object
  file_name <- basename(file_path)
  file_name <- sub("\\.txt\\.gz$", "", file_name)
  
  # Create a Seurat object
  seurat_obj <- CreateSeuratObject(counts = data, project = file_name)
  
  # Store the Seurat object in the list
  seurat_objects[[file_name]] <- seurat_obj
}

# Optionally, print out the Seurat objects
print(seurat_objects)

```
##Normalize the matrix find highly variable genes and scale the data

```{r}
# Separate loop to normalize data for each Seurat object
for (file_name in names(seurat_objects)) {
  seurat_objects[[file_name]] <- NormalizeData(seurat_objects[[file_name]])
}

# Separate loop to find variable features for each Seurat object
for (file_name in names(seurat_objects)) {
  seurat_objects[[file_name]] <- FindVariableFeatures(seurat_objects[[file_name]])
}

# Separate loop to scale data for each Seurat object
for (file_name in names(seurat_objects)) {
  seurat_objects[[file_name]] <- ScaleData(seurat_objects[[file_name]])
}

```

##PCA plot

```{r}
for (file_name in names(seurat_objects)) {
  # Perform PCA
  seurat_objects[[file_name]] <- RunPCA(seurat_objects[[file_name]])
  
  # Visualize PCA (using the first 2 dimensions)
  pca_plot <- DimPlot(seurat_objects[[file_name]], reduction = "pca", dims = c(1, 2)) + 
    ggtitle(paste(file_name, "- PCA Plot"))
  
  # Print the PCA plot (optional: save it as a file if needed)
  print(pca_plot)
}

```
##Cluster the cells and perform UMAP 

```{r}
# Separate loop to perform clustering and UMAP visualization for each Seurat object
for (file_name in names(seurat_objects)) {
  # Find neighbors using the first 10 PCs
  seurat_objects[[file_name]] <- FindNeighbors(seurat_objects[[file_name]], dims = 1:10)
  
  # Perform clustering
  seurat_objects[[file_name]] <- FindClusters(seurat_objects[[file_name]], resolution = 0.5)  # Adjust resolution as needed
  
  # Run UMAP for dimensionality reduction
  seurat_objects[[file_name]] <- RunUMAP(seurat_objects[[file_name]], dims = 1:10)
  
  # Visualize the clusters on UMAP
  umap_plot <- DimPlot(seurat_objects[[file_name]], reduction = "umap", group.by = "seurat_clusters") + 
    ggtitle(paste(file_name, "- UMAP with Clusters"))
  
  # Print the UMAP plot (optional: save it as a file if needed)
  print(umap_plot)
}

```

